Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-03-17 paper/manuscript.Rmd
intro edits

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between Western and Eastern Indian
Islands (Figure 1), a distance of at least 470 m which took ~9 minutes (~52 m per
minute). This observation prompted us to investigate whether swimming behaviour
was common among GPS radio-collared caribou in the population (n = 29 adult
females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada). We
identified swimming events as two consecutive GPS (2-hour relocation rates)
locations from an individual occurred on different islands. Northeastern
Newfoundland typically experiences pack ice during winter and caribou are known
locally to travel between islands by walking over the ice. We restricted our GPS
data to the ice-free period of the year (April 1 - November 25)<!--confirm dates
--> and we identified three adult female caribou that collectively had 86
swimming events over three years (Figure 2). In addition to our own visual and
remotely sensed observations, residents of Fogo Island have also reported
observing caribou swimming between islands on numerous occasions.

@@

Predation pressure is also known to affect caribou swimming behaviour, both in
immediate situations to escape predators [@Bergerud1990] and also pre-emptively
to give birth or raise calves in safer areas like isolated islands
[@Bergerud1985]. The Fogo Island archipelago does not contain historic caribou
predators such as wolves (*Canis lupus*) or black bears (*Ursus americanus*),
leaving coyotes (*Canis latrans*) as the top predator. Coyotes may predate
calves [@Rayl2014] but do not pose a substantial risk to adult caribou
[@Bastille-Rousseau2016]; predation in general does not exert as strong a force
on Fogo Island caribou as in mainland populations. Calves that are most
vulnerable to predation likely face a much higher energetic cost of oceanic
swimming than do adults, and indeed none of the females observed swimming had a
calf with them at the time.<!-- female-centric rather than time-of-year
["swimming events occurred in the weeks after calves were born"] seems better,
if we can confirm this--> Given the frequency of these swimming events, and that
swimming in the ocean with a calf-at-heel is riskier than other predator
avoidance strategies, we suspect that predation risk is not a main driver of
oceanic swimming in Fogo Island caribou.

Forage limitation and over-grazing is a major concern for caribou populations
[@Schaefer2016] and could explain movement between islands. Conspecific
competition for resources driven by reduced foraging opportunities can decrease
reproductive success of adult females [@Schaefer2016].Forage scarcity has been
proposed as a potential reason for caribou moving between arctic islands on the
sea-ice in winter [@Miller1977]. Caribou were introduced to Fogo Island (n = 26
animals between 1964–67) and Change Island (n = 5 animals in 1964) from
Newfoundland as part of a series of translocations and introductions throughout
the province (Bergerud & Mercer 1989). It is unknown when caribou began
colonizing nearby islands, but given that some are small, it is likely forage is
depleted relatively quickly. While forage depletion is possible for Fogo Island,
For very small islands, over-grazing by newly arrived caribou may rapidly
deplete forage [@Bergerud1990].

>>>>

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between Western and Eastern Indian
Islands (Figure 1), a distance of at least 470 m which took ~9 minutes (~52 m per
minute). This observation prompted us to investigate whether swimming behaviour
was common among GPS radio-collared caribou in the population (n = 29 adult
females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada). We
identified swimming events as two consecutive GPS (2-hour relocation rates)
locations from an individual occurred on different islands. Northeastern
Newfoundland typically experiences pack ice during winter and caribou are known
locally to travel between islands by walking over the ice. We restricted our GPS
data to the ice-free period of the year (April 1 - December 31) <!--confirm dates
--> and we identified three adult female caribou that collectively had 86
swimming events over three years (Figure 2). In addition to our own visual and
remotely sensed observations, residents of Fogo Island have also reported
observing caribou swimming between islands on numerous occasions.

@@

<!-- DO WE WANT TO CUT THIS? -->
Predation pressure is also known to affect caribou swimming behaviour, both in
immediate situations to escape predators [@Bergerud1990] and also pre-emptively
to give birth or raise calves in safer areas like isolated islands
[@Bergerud1985]. The Fogo Island archipelago does not contain historic caribou
predators such as wolves (*Canis lupus*) or black bears (*Ursus americanus*),
leaving coyotes (*Canis latrans*) as the top predator. Coyotes may predate
calves [@Rayl2014] but do not pose a substantial risk to adult caribou
[@Bastille-Rousseau2016]; predation in general does not exert as strong a force
on Fogo Island caribou as in mainland populations. Calves that are most
vulnerable to predation likely face a much higher energetic cost of oceanic
swimming than do adults, and indeed none of the females observed swimming had a
calf with them at the time.<!-- female-centric rather than time-of-year
["swimming events occurred in the weeks after calves were born"] seems better,
if we can confirm this--> Given the frequency of these swimming events, and that
swimming in the ocean with a calf-at-heel is riskier than other predator
avoidance strategies, we suspect that predation risk is not a main driver of
oceanic swimming in Fogo Island caribou.

Forage limitation and over-grazing is a major concern for caribou populations
[@Schaefer2016] and could explain movement between islands. Conspecific
competition for resources driven by reduced foraging opportunities can decrease
reproductive success of adult females [@Schaefer2016].Forage scarcity has been
proposed as a potential reason for caribou moving between arctic islands on the
sea-ice in winter [@Miller1977]. Caribou were introduced to Fogo Island (n = 26
animals between 1964–67) and Change Island (n = 5 animals in 1964) from
Newfoundland as part of a series of translocations and introductions throughout
the province (Bergerud & Mercer 1989). It is unknown when caribou began
colonizing nearby islands, but given that some are small, over-grazing by newly
arrived caribou may rapidly deplete forage [@Bergerud1990]. We posit that some caribou swim to nearby islands and forage
